---
title: "Data-Visualization"
author: "Rosie Wu"
date: "2024-11-11"
output: html_document
---
# How temperature, energy load, and electricity price changed and correlated over year 2016-2023
# Prediction of number of extreme cold days in 2024-2025 and its required energy load based on different scenarios

```{r setup, include=FALSE}
#install.packages("httr")
#install.packages("jsonlite")
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(readr)
library(here)
```
# Data import and wrangling for correlation and visualization

```{r}
# This visualizes the temperature change overtime (year 2016-2023)
# Load necessary libraries
library(ggplot2)
temp_data <- read.csv("Data/Processed/Temperature_MaxMin_Daily.csv")

# Ensure the date column is in Date format
temp_data$Date <- as.Date(temp_data$Date)

# Aggregate the data to monthly averages
monthly_min_temp <- temp_data %>%
  mutate(month = format(Date, "%Y-%m")) %>% # Extract year-month
  group_by(month) %>%
  summarise(
    avg_min_temp = mean(Temperature_Min, na.rm = TRUE)
  ) %>%
  ungroup()

# Convert the month column to Date for plotting
# I assigned 01 to Days in the Dates, but it's actually average done earlier
min_temp_data$month <- as.Date(paste0(min_temp_data$month, "-01"))

# Ensure the data is numeric
min_temp_data$avg_min_temp <- as.numeric(as.character(min_temp_data$avg_min_temp))

```


```{r}
# get Energy Load Data
EnergyLoad_Daily <- read_csv("Data/Processed/EnergyLoad_Daily.csv")
View(EnergyLoad_Daily)

# Ensure the date column is in Date format
EnergyLoad_Daily$Date <- as.Date(EnergyLoad_Daily$Date)

# Aggregate the data to monthly averages
monthly_load <- EnergyLoad_Daily %>%
  mutate(month = format(Date, "%Y-%m")) %>% # Extract year-month
  group_by(month) %>%
  summarise(
    MW = mean(`ERCOT(MW)`, na.rm = TRUE)
  ) %>%
  ungroup()

# Convert the month column to Date for plotting
# I assigned 01 to Days in the Dates, but it's actually average done earlier
monthly_load$month <- as.Date(paste0(monthly_load$month, "-01"))

# Ensure the data is numeric
monthly_load$MW <- as.numeric(as.character(monthly_load$MW))

```

```{r}
# import data for electricity price 
library(readr)
ElectricityPrice_Monthly <- read_csv("Data/Processed/ElectricityPrice_Monthly.csv")
View(ElectricityPrice_Monthly)

# the Price data is in cent/kwh, so I need to convert the Energy load (MW) previously to KW
# Add a new column to energy_load_data for kW
monthly_load$load_kw <- monthly_load$MW * 1000

```

```{r}
# rename the column name to combine the datasets
ElectricityPrice_Monthly <- ElectricityPrice_Monthly %>%
  rename(month = Date)

# Merge the datasets by the 'month' column (ensure all have the same 'month' column)
combined_data <- min_temp_data %>%
  left_join(monthly_load, by = "month") %>%
  left_join(ElectricityPrice_Monthly, by = "month")

# rename the column name of price in combined datasets since it had paranthesis and difficult to work with
combined_data <- combined_data %>% rename(Price = 'Price(cent/KWh)')

```

# Correlation and Plot the correlation between energy load and electricity price

```{r}
# Calculate the correlation between energy load (in kW) and electricity price (in cents/kWh)
correlation <- cor(combined_data$load_kw, combined_data$Price, use = "complete.obs")

# Print the correlation coefficient
print(paste("Correlation coefficient:", correlation))

# Plot the correlation between energy load and electricity price
ggplot(combined_data, aes(x = load_kw, y = Price)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Energy Load and Electricity Price",
    x = "Energy Load (kW)",
    y = "Electricity Price (cents/kWh)"
  ) +
  theme_minimal()

```

# Correlation and Plot the correlation between Min Temperature and Energy Load

```{r}
# Calculate the correlation between energy load (in kW) and Monthly Avg temperature
correlation_2 <- cor(combined_data$avg_min_temp, combined_data$load_kw, use = "complete.obs")

# Print the correlation coefficient
print(paste("Correlation coefficient:", correlation_2))

# Plot the correlation between energy load and electricity price
ggplot(combined_data, aes(x = avg_min_temp, y = load_kw)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Monthly Minimum Temperature & Energy Load",
    x = "Monthly Average Min Temperature",
    y = "Energy Load (kW)"
  ) +
  theme_minimal()

```

# Correlation and Plot the correlation between Min Temperature and Electricity Price

```{r}
# Calculate the correlation between monthly min temperature & electricity price 
correlation_3 <- cor(combined_data$avg_min_temp, combined_data$Price, use = "complete.obs")
print(paste("Correlation coefficient:", correlation_3))

ggplot(combined_data, aes(x = avg_min_temp, y = Price)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Energy Load and Electricity Price",
    x = "Monthly Average Min Temperature",
    y = "Electricity Price (cents/kWh)"
  ) +
  theme_minimal()

```

