```{r}
# Set up r
library(tidyverse)
library(lubridate)
install.packages("trend")
library(trend)
install.packages("zoo")
library(zoo)
install.packages("Kendall")
library(Kendall)
install.packages("tseries")
library(tseries)
library(ggplot2)
# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```


```{r}
# Import data
ElectricityPrice_Monthly <-
  read.csv("./Data/Processed/ElectricityPrice_Monthly.csv")
EnergyLoad_Daily <-
  read.csv("./Data/Processed/EnergyLoad_Daily.csv")
Temperature_Daily <-
  read.csv("./Data/Processed/Temperature_MaxMin_Daily.csv")
```

Generate three time series objects based on imported data
```{r}
electricity.monthly.ts <- ts(ElectricityPrice_Monthly$Price.cent.KWh.,
                   start=c(2016,01), end = c(2023,12),
                   frequency=12) 
energyload.daily.ts <- ts(EnergyLoad_Daily$ERCOT.MW.,
                   start=c(2016,01,01), end = c(2023,12,31),
                   frequency=365)
temp.min.daily.ts <- ts(Temperature_Daily$Temperature_Min,
                   start=c(2016,01,01), end = c(2023,12,31),
                   frequency=365)
```

Decompose the time series objects and plot them.
```{r}
energyload.daily.decomp <- stl(energyload.daily.ts,s.window = "periodic")
temp.min.daily.decomp <- stl(temp.min.daily.ts,s.window = "periodic")
electricity.monthly.decomp <- stl(electricity.monthly.ts,s.window = "periodic")

temp.ts <- plot(temp.min.daily.decomp)
energy.ts <- plot(energyload.daily.decomp)
electricity.ts <- plot(electricity.monthly.decomp)
 
```

Do a seasonal Mann Kendall test to evaluate trends in the three datasets
```{r}
electricity.monthly.trend <- Kendall::SeasonalMannKendall(electricity.monthly.ts)
energyload.daily.trend <- Kendall::SeasonalMannKendall(energyload.daily.ts)
temp.min.daily.trend <- Kendall::SeasonalMannKendall(temp.min.daily.ts)

summary(electricity.monthly.trend)

summary(energyload.daily.trend)

summary(temp.min.daily.trend)
```
The seasonal Mann Kendall tests showed that electricity price and energy load has been statistically increasing throughout 2016 to 2023; however, min temperature has been decreasing, but not statistically significant.
