---
title: "Time Series Analysis"
author: "Judy Zhu"
date: "2024-12-11"
output: html_document
---
  
```{r}
# Set up r
library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)
library(ggplot2)
# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)
```


```{r}
# Import data
ElectricityPrice_Monthly <-
  read.csv("./Data/Processed/ElectricityPrice_Monthly.csv")
EnergyLoad_Daily <-
  read.csv("./Data/Processed/EnergyLoad_Daily.csv")
EnergyLoad_DailyAvg <-
  read.csv("./Data/Processed/Monthly_Load.csv")
Temperature_Daily <-
  read.csv("./Data/Processed/Temperature_MaxMin_Daily.csv")

```

Generate three time series objects based on imported data
```{r}
electricity.monthly.ts <- ts(ElectricityPrice_Monthly$Price.cent.KWh.,
                   start=c(2016,01), end = c(2023,12),
                   frequency=12) 
energyload.daily.ts <- ts(EnergyLoad_Daily$ERCOT.MW.,
                   start=c(2016,01,01), end = c(2023,12,31),
                   frequency=365)
energyload.daily.avg.ts <- ts(EnergyLoad_DailyAvg$MW,
                   start=c(2016,01), end = c(2023,12),
                   frequency=12)
temp.min.daily.ts <- ts(Temperature_Daily$Temperature_Min,
                   start=c(2016,01,01), end = c(2023,12,31),
                   frequency=365)

```

Decompose the time series objects and plot them.
```{r}
energyload.daily.decomp <- stl(energyload.daily.ts,s.window = "periodic")
energyload.daily.avg.decomp <- stl(energyload.daily.avg.ts,s.window = "periodic")
temp.min.daily.decomp <- stl(temp.min.daily.ts,s.window = "periodic")
electricity.monthly.decomp <- stl(electricity.monthly.ts,s.window = "periodic")

temp.ts <- plot(temp.min.daily.decomp)
energy.ts <- plot(energyload.daily.decomp)
energy.avg.ts <- plot(energyload.daily.avg.decomp)
electricity.ts <- plot(electricity.monthly.decomp)

```

Do a seasonal Mann Kendall test to evaluate trends in the three datasets.
```{r}
electricity.monthly.trend <- Kendall::SeasonalMannKendall(electricity.monthly.ts)
energyload.daily.trend <- Kendall::SeasonalMannKendall(energyload.daily.ts)
temp.min.daily.trend <- Kendall::SeasonalMannKendall(temp.min.daily.ts)

summary(electricity.monthly.trend)
summary(energyload.daily.trend)
summary(temp.min.daily.trend)
```
The seasonal Mann Kendall tests results showed that electricity price and energy load has been statistically increasing throughout 2016 to 2023 (positive score with p-value < 0.05); however, minimal temperature has been decreasing, but not statistically significant (p-value =0.72513 > 0.05) . We use seasonal Mann Kendall test because seasonal pattern is significant in these three datasets. Since electricity price is influenced by multiple factors, such as market dynamics and policies, we decide to run a Mann Kendall test on detrended electricity data.
```{r}
electricity.monthly.Components <- 
  as.data.frame(electricity.monthly.decomp$time.series[,])
electricity.monthly.detrended <- 
  ElectricityPrice_Monthly$Price.cent.KWh. - electricity.monthly.Components$seasonal

electricity.monthly.detrended.result <- MannKendall(electricity.monthly.detrended)
summary(electricity.monthly.detrended.result)
```

Detrended Mann Kendall test's result further supports that electricity price has been statistically significantly increasing throughout the past 8 years.

Lastly, we hope to predict future energy load needed based on historical data. We will use Autoregressive Integrated Moving Average (ARIMA) package to do a simple forecasting using historical energy load data from 2016 to 2023. Daily energy load data will contain too much noise and excessive workload for the function to run; therefore, we choose to use averaged daily load of each months.
```{r}
library(forecast)
energy.prediction <- auto.arima(energyload.daily.avg.ts, seasonal = TRUE)
summary(energy.prediction)
# Check residuals
checkresiduals(energy.prediction)
# Forecast
energy.forecasted <- forecast(energy.prediction, h = 24)
plot(energy.forecasted)
summary(energy.forecasted)
```






