---
title: "Data-Visualization"
author: "Rosie Wu"
date: "2024-11-11"
output: html_document
---
# How temperature, energy load, and electricity price changed and correlated over year 2016-2023
# Prediction of number of extreme cold days in 2024-2025 and its required energy load based on different scenarios

```{r setup, include=FALSE}
#install.packages("httr")
#install.packages("jsonlite")
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
```
# Data import and wrangling for correlation and visualization

```{r}
# This visualizes the temperature change overtime (year 2016-2023)
temp_data <- read.csv("Data/Processed/Temperature_MaxMin_Daily.csv")

# Ensure the date column is in Date format
temp_data$Date <- as.Date(temp_data$Date)

# import monthly electricity load data
Monthly_Load <- read.csv("./Data/Processed/Monthly_Load.csv")
Monthly_Load$month <- as.Date(paste0(Monthly_Load$month, "-01"), format = "%Y-%m-%d")

# import data for electricity price 
library(readr)
ElectricityPrice_Monthly <- read_csv("Data/Processed/ElectricityPrice_Monthly.csv")

# the Price data is in cent/kwh, so I need to convert the Energy load (MW) previously to KW
# Add a new column to energy_load_data for kW
Monthly_Load$load_kw <- Monthly_Load$MW * 1000

# rename the column name to combine the datasets
ElectricityPrice_Monthly <- ElectricityPrice_Monthly %>%
  rename(month = Date)
```

# Since we changed our Temperature monthly to # of days in extreme low temperature (<=0 celsius), so now we add a new column for this # of cold days, and combine an updated dataset.

```{r}
#Build a new column "Extreme_Cold"
temp_data <- temp_data %>%
  mutate(Extreme_Cold = ifelse(Temperature_Min <= 0, 1, 0))

# Aggregate the number of extreme cold days by month
extreme_cold_monthly <- temp_data %>%
  mutate(month = floor_date(Date, "month")) %>%  # Extract the month
  group_by(month) %>%  # Group by month
  summarise(`Extreme_Cold_days` = sum(Extreme_Cold))

# combine all datasets together
combined_data <- Monthly_Load %>%
  left_join(extreme_cold_monthly, by = "month", na.rm==T) %>%
  left_join(ElectricityPrice_Monthly, by = "month", na.rm==T)

# rename the column name of price in combined datasets since it had paranthesis and difficult to work with
combined_data <- combined_data %>% rename(Price = 'Price(cent/KWh)')

# the last row of dataset is 2024-01, but the data is still not collected completely, so eliminate this row to ensure accuracy
combined_data <- combined_data[-nrow(combined_data), ]
```


# Exploratory visualizations for each variable to be shown before all of the correlations and Time Series analysis.

# Below is a column graph of all of the months' # of days below 0 degree
```{r}
ggplot(combined_data, aes(x = month, y = Extreme_Cold_days)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Number of Days Below 0°C Per Month (2016–2023)",
    x = "Month",
    y = "Number of Days"
  ) +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "6 months") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
> According to the visualization of the # of days <0 degree celsius in each month in Texas from 2016-01 to 2023-12, number of below-freezing days in 2021-01 and 2022-01 are the highest. Also, the time range (number of months) for extremely cold days are highest in 2021, which align with the news of power breakouts in 2021 in Texas due to the abnormal cold-front.


# Below is a line of graph for Monthly load from 2016-2023

```{r}
ggplot(combined_data, aes(x = month, y = MW)) +
  geom_line(color = "blue", size = 1) +  # Line graph
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Monthly Energy Load in MW in Texas in 2016-2023",
    x = "Month",
    y = "Energy Load (MW)")

```
> According to the line graph from 2016 to 2023, the electricity load is generally increasing overtime with great pattern of seasonality, and it has spiked in mid-2023. 



# Below is a line graph for monthly electricity price from 2016-2023
```{r}
ggplot(combined_data, aes(x = month, y = Price)) +
  geom_line(color = "blue", size = 1) +  # Line graph
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Monthly Average Electricity Price in Texas 2016-2023",
    x = "Month",
    y = "Electricity Price (Cent/KWh)")

```
> The electricity price in Texas is increasing overtime from 2016-2023 with slight but less consistent seasonality. However, the electricity price spike around early 2021, which could be a reflection from the large-scale blackout in the winter in Texas. The state-wide grid and local distribution systems were struggling to recover to cater the demand from the local residents and business during and right after the cold-front.



# Correlation and Plot the correlation between energy load and electricity price

```{r}
# Calculate the correlation between energy load (in kW) and electricity price (in cents/kWh)
correlation <- cor(combined_data$load_kw, combined_data$Price, use = "complete.obs")

# Print the correlation coefficient
print(paste("Correlation coefficient:", correlation))

# Plot the correlation between energy load and electricity price
ggplot(combined_data, aes(x = load_kw, y = Price)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Energy Load and Electricity Price",
    x = "Energy Load (kW)",
    y = "Electricity Price (cents/kWh)"
  ) +
  theme_minimal()

```

# Correlation and Plot the correlation between Min Temperature and Energy Load

```{r}
# Calculate the correlation between energy load (in kW) and Monthly Avg temperature
correlation_2 <- cor(combined_data$avg_min_temp, combined_data$load_kw, use = "complete.obs")

# Print the correlation coefficient
print(paste("Correlation coefficient:", correlation_2))

# Plot the correlation between energy load and electricity price
ggplot(combined_data, aes(x = avg_min_temp, y = load_kw)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Monthly Minimum Temperature & Energy Load",
    x = "Monthly Average Min Temperature",
    y = "Energy Load (kW)"
  ) +
  theme_minimal()

```

# Correlation and Plot the correlation between Min Temperature and Electricity Price

```{r}
# Calculate the correlation between monthly min temperature & electricity price 
correlation_3 <- cor(combined_data$avg_min_temp, combined_data$Price, use = "complete.obs")
print(paste("Correlation coefficient:", correlation_3))

ggplot(combined_data, aes(x = avg_min_temp, y = Price)) +
  geom_point(color = "blue", size = 2) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Add regression line
  labs(
    title = "Correlation Between Minimimum Temperature and Electricity Price",
    x = "Monthly Average Min Temperature",
    y = "Electricity Price (cents/kWh)"
  ) +
  theme_minimal()

```




